---
title: "USArrests"
author: "Gale, Amanda"
date: "2025-09-16"
output: html_document
---

# Task I
```{r}
df <- USArrests
```


```{r}
str(df)
```

## Find Outliers
```{r}
av.murders <- mean(df$Murder)
sd.murders <- sd(df$Murder)
ind <- which((df$Murder > 1.5*sd.murders + av.murders) | 
               (df$Murder < av.murders - 1.5*sd.murders)) 
print("outlier states:" + df[ind,0])
```


## Evaluate Normality
```{r}
shapiro.test(df$Murder)
shapiro.test(df$UrbanPop)
```

```{r}
qqnorm(df$Murder)
qqline(df$Murder)
```


```{r}
qqnorm(df$UrbanPop)
qqline(df$UrbanPop)
```

## Evaluate Correlation of Murders vs Population
```{r}
# based on assumption of normality, pearson test
cor(df$UrbanPop, df$Murder)
```

```{r}
plot(df$UrbanPop, df$Murder)
```

Based on the Pearson Correlation test value of close to 0 and the visual results of the plot, we can conclude that murder rates are not correlationed to urban population.


# Task II
Based on the data on the growth of mobile phone use in Brazil (you’ll need to copy the data and create a CSV that you can load into R or use the gsheet2tbl() function from the gsheet package), forecast phone use for the next time period using a 2-year weighted moving average (with weights of 5 for the most recent year, and 2 for other), exponential smoothing (alpha of 0.4), and linear regression trendline.

```{r}
library(gsheet)
library(forecast)

url <- "https://docs.google.com/spreadsheets/d/1tOnM9XceK4Ak8tzWQ2vDelWlJexzJiS3LbT6MN6_rW0/edit?usp=sharing"
df <- read.csv(text=gsheet2text(url, format='csv'), 
               stringsAsFactors=FALSE,
               encoding = "UTF-8")
```

### Visualize Data
```{r}
plot(df$Subscribers)
```

## Predict Month 12 Subscribers

### WMA
```{r}
weights <- c(2,5)
wma <- TTR::WMA(df$Subscribers[1:11], n=length(weights), wts=weights)
wma.forecast <- round(sum(wma[c(10,11)] * weights)/7)
wma.forecast
wma[12] <- wma.forecast
wma <- c(NA, wma)
wma
```

### SES
```{r}
# manually creating the loop
a <- 0.4
df$Ft <- 0
df$E <- 0

df$Ft[1] <- df$Subscribers[1]
df$E[1] <- 0

for (i in 2:nrow(df)) {
  df$Ft[i] <- df$Ft[i-1] + a*df$E[i-1]
  df$E[i] <- df[i,2]  - df$Ft[i] 
}

ses.forecast <- round(df$Ft[12])
ses.forecast
```

```{r}
# built in function 
df.ts <- ts(df$Subscribers,
            start = 1,
            frequency = 12)

ses.model <- HoltWinters(df.ts[1:11],
                         alpha=0.4,
                         beta=F,
                         gamma=F)

ses.forecast2 <- forecast(ses.model, h=1)
ses.forecast2[4]
```

### Linear Regression
```{r}
lm.model <- lm(df$Subscribers~df$Year)

# calculation over time using predict() function
lm.forecast <- predict(lm.model, newdata=data.frame(period=12))

lm.forecast[12]

# manual calculation using coefficients
lm.trendline <- function(x) {
  lm.model$coefficients[2]*x + lm.model$coefficients[1]
}

lm.trendline(12)

plot(lm.trendline(seq(1:12)))
```


## MSE Calculations
Subtract predicted subscribers from actual for each model.

```{r}
df.mse <- data.frame(
 WMA=0,
 SES=0,
 LinReg=0
  )
# WMA
df$wma.error <- df$Subscribers - wma
df.mse$WMA <- mean(df$wma.error^2, na.rm = TRUE)

# SES
df$ses.error <- df$Subscribers - df$Ft
df.mse$SES <- mean(df$ses.error^2, na.rm = TRUE)

# LM
df$lm.error <- df$Subscribers - lm.forecast
df.mse$LinReg <- mean(df$lm.error^2, na.rm = TRUE)

df.mse
```



```{r}

```



